---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "fig/README-"
)
```

# polypoly

The goal of polypoly is to provide helper functions for polynomials created by 
`poly()`.

```{r logo, message = FALSE, echo = FALSE, fig.asp=.2, dpi=600, fig.width = 8}
library(ggplot2)
x <- poly(1:10000, degree = 6)
df <- polypoly::poly_melt(x)
df$dg <- as.numeric(df$degree)
ggplot(df) + 
  aes(x = observation, y = value, color = dg, group = dg) + 
  geom_line(size = 1) + 
  viridis::scale_color_viridis(option = "viridis") + 
  theme_void() +
  guides(color = "none") +
  coord_cartesian(xlim = NULL, ylim = NULL, expand = FALSE)
```

## Installation

You can install polypoly from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("tjmahr/polypoly")
```

## Examples

This package provides a tidying function `poly_melt()` and plotting function
`poly_plot()`.

```{r example, fig.width = 5, fig.height = 3}
library(polypoly)
xs <- 1:40
poly_mat <- poly(xs, degree = 5)
poly_melt(poly_mat)
poly_plot(poly_mat)
```

We can also plot raw polynomials, but that display is less useful because the
_x_-axis corresponds to the row number of polynomial matrix.

```{r raw-example, fig.width = 5, fig.height = 3}
poly_raw_mat  <- poly(-10:10, degree = 3, raw = TRUE)
poly_plot(poly_raw_mat)
```

We can make the units clearer by using `by_observation = FALSE` so that the
_x_-axis corresponds to the first column of the polynomial matrix.

```{r raw-by-degree1,  fig.width = 5, fig.height = 3}
poly_raw_mat  <- poly(-10:10, degree = 3, raw = TRUE)
poly_plot(poly_raw_mat, by_observation = FALSE)
```


## Resources

If you searched for help on `poly()`, see also:

* [What does the R function `poly()` really do?](https://stackoverflow.com/questions/19484053/what-does-the-r-function-poly-really-do)
* [Source code for the poly function](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/library/stats/R/contr.poly.R#L85)
