---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "fig/README-"
)
```

# polypoly

The goal of polypoly is to provide helper functions for polynomials created by
`poly()`. 

The name "polypoly" has no special significance It's just something catchy and
related to `poly()`.

## Installation

You can install polypoly from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("tjmahr/polypoly")
```

## Example

This package provides a tidying function `poly_melt()` and plotting function
`poly_plot()`.


```{r example, fig.width = 5, fig.height = 3}
library(polypoly)
xs <- 1:40
poly_mat <- poly(xs, degree = 5)
poly_melt(poly_mat)
poly_plot(poly_mat)
```

We can also plot raw polynomials, but that display is less useful because the
x-axis corresponds to the row number of polynomial matrix.

```{r raw-example, fig.width = 5, fig.height = 3}
poly_raw_mat  <- poly(-10:10, degree = 3, raw = TRUE)
poly_plot(poly_raw_mat)
```



## Resources

If you searched for help on `poly()`, see also:

* [What does the R function `poly()` really do?](https://stackoverflow.com/questions/19484053/what-does-the-r-function-poly-really-do)
