---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "fig/README-"
)
```


```{r logo, message = FALSE, echo = FALSE, fig.asp=.2, dpi=600, fig.width = 8}
library(ggplot2)
x <- poly(1:10000, degree = 6)
df <- polypoly::poly_melt(x)
df$dg <- as.numeric(df$degree)
ggplot(df) + 
  aes(x = observation, y = value, color = dg, group = dg) + 
  geom_line(size = 1) + 
  viridis::scale_color_viridis(option = "viridis") + 
  theme_void() +
  guides(color = "none") +
  coord_cartesian(xlim = NULL, ylim = NULL, expand = FALSE)
```

# polypoly

Helper functions for polynomials created by `poly()`.

## Installation

You can install polypoly from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("tjmahr/polypoly")
```

## Examples

### Tidying

This package provides a tidying function `poly_melt()`.

```{r}
library(polypoly)
xs <- 1:40
poly_mat <- poly(xs, degree = 5)
poly_melt(poly_mat)
```


### Plotting

Plot a matrix with `poly_plot()`.

```{r example, fig.width = 5, fig.height = 3}
poly_plot(poly_mat)
```

We can also plot raw polynomials, but that display is less useful because the
_x_-axis corresponds to the row number of polynomial matrix.

```{r raw-example, fig.width = 5, fig.height = 3}
poly_raw_mat  <- poly(-10:10, degree = 3, raw = TRUE)
poly_plot(poly_raw_mat)
```

We can make the units clearer by using `by_observation = FALSE` so that the
_x_-axis corresponds to the first column of the polynomial matrix.

```{r raw-by-degree1,  fig.width = 5, fig.height = 3}
poly_raw_mat  <- poly(-10:10, degree = 3, raw = TRUE)
poly_plot(poly_raw_mat, by_observation = FALSE)
```

### Rescaling a matrix

The ranges of the terms created by `poly()` are sensitive to repeated values.

```{r}
p1 <- poly(0:9, degree = 2)
p2 <- poly(rep(0:9, 18), degree = 2)

col_range <- function(matrix) {
  apply(matrix, 2, function(xs) max(xs) - min(xs))  
}

col_range(p1)
col_range(p2)
```

Thus, two models fit with `y ~ poly(x, 3)` will not have comparable coefficients
when the number of rows changes, even if the unique values of `x` did not
change!

`poly_rescale()` adjusts the values in the polynomial matrix so that the linear
component has a specified range. The other terms are scaled by the same factor. 

```{r rescaled,  fig.width = 5, fig.height = 3}
col_range(poly_rescale(p1, scale_width = 1))
col_range(poly_rescale(p2, scale_width = 1))

poly_plot(poly_rescale(p2, scale_width = 1), by_observation = FALSE)
```

### Adding columns to a dataframe

`poly_add_columns()` adds orthogonal polynomial transformations of a predictor
variable to a dataframe.

Here's how we could add polynomials to the `sleepstudy` dataset.

```{r}
df <- tibble::as_tibble(lme4::sleepstudy)
print(df)

poly_add_columns(df, Days, degree = 3)
```

We can optionally customize the column names and rescale the polynomial terms.

```{r}
poly_add_columns(df, Days, degree = 3, prefix = "poly_", scale_width = 1)
```

We can confirm that the added columns are orthogonal.

```{r sleepstudy, fig.width = 5, fig.height = 3}
df2 <- poly_add_columns(df, Days, degree = 3, scale_width = 1)

dfl <- tidyr::gather(df2, var, value, Days1, Days2, Days3)
ggplot(dfl) + 
  aes(x = Days, y = value, color = var) + 
  geom_line()
```




## Resources

If you searched for help on `poly()`, see also:

* [What does the R function `poly()` really do?](https://stackoverflow.com/questions/19484053/what-does-the-r-function-poly-really-do)
* [Source code for the poly function](https://github.com/wch/r-source/blob/af7f52f70101960861e5d995d3a4bec010bc89e6/src/library/stats/R/contr.poly.R#L85)
